---
title: "Rockwood index outperforms patient age as a predictor of poor patient
        outcomes."
author: 'Alex Bokov ^1,✉^, Sara Espinoza ^1^, Chandana Tripathy ^1^,
  ..., ..., ..., and Kathleen R. Stevens ^1^'
css: "production.css"
abstract: 'We have found that the electronic frailty index (EFI), a risk
  score developed using the Rockwood deficit-accumulation framework, is
  a strong predictor of falls, hospitalizations, hospital-acquired
  infections, and loss of independence without relying on any predictors
  including patient age. In fact, EFI outperformed patient age as a
  predictor. EFI''s accuracy was not negated by the fact that we used
  patients of all ages, not just older patients.'
documentclass: article
clean: false
bibliography: efi_paper.bib
csl: harvard-cite-them-right.csl
output:
  bookdown::html_document2:
    fig_caption: true
    self_contained: true
    number_sections: false
    keep_md: true
    clean: false
  bookdown::word_document2:
    reference_docx: efi_template.docx
    fig_caption: true
    self_contained: true
    number_sections: false
---

```{r load_deps, echo=FALSE, message=FALSE, warning=FALSE,results='hide'}
.projpackages <- c('GGally','pander','dplyr','ggplot2','data.table'
                   ,'survival','broom','forcats','table1','english');
.deps <- c( 'analysis.R' );
.debug <- 0;
.junk<-capture.output(source('./scripts/global.R',chdir=TRUE,echo=FALSE
                             ,local=TRUE));
# Set some formatting options for this document
panderOptions('table.alignment.default','right');
panderOptions('table.alignment.rownames','right');
panderOptions('table.split.table',Inf);
panderOptions('p.wrap','');
panderOptions('p.copula',', and ');
panderOptions('graph.fontfamily','serif');
# theme_get() gives default theme
theme_set(theme_bw(base_family = 'serif',base_size=14) +
            theme(strip.background = element_rect(fill=NA,color=NA)
                  ,strip.text = element_text(size=15)));
knitr::opts_chunk$set(echo=.debug>0, warning=.debug>0, message=.debug>0);

# detect current output format
.outfmt <- knitr::opts_knit$get('rmarkdown.pandoc.to');
message('.outfmt = ',.outfmt);
if(is.null(.outfmt)){
  if(knitr::is_html_output()) .outfmt <- 'html' else {
    if(knitr::is_latex_output()) .outfmt <- 'latex' else {
      .outfmt <- 'unknown';
    }}};
message('.outfmt = ',.outfmt);

.currentscript <- current_scriptname('efi_safety.Rmd');
.outcomenames <- c('vi_c_psi','vi_c_hospinfect','vi_c_hosptrauma'
                   ,'vi_c_cardiac');

# limit the data to items of interest
tb2 <- subset(tb2,outcome %in% .outcomenames & predictor %in%
                c('Frailty','Frailty, age:18-45','Frailty, age:45-65'
                  ,'Frailty, age:65+'));
tb3 <- subset(tb3,Predictor %in% c('Frailty','Frailty, age:18-45'
                                   ,'Frailty, age:45-65','Frailty, age:65+'));
fits <- fits[.outcomenames];
snip_resultsfold <- unique(resultsfold00(tb2));
```


# Introduction

Frailty is the lifelong erosion of stress resistance and accumulation of
impairments across multiple physiological systems. Among older community
dwelling adults 32% have been classified as pre-frail and 24% have been
classified as frail [@hoover13]. Frailty predicts disability,
injurious falls, and mortality [@pajewski19], emergency room visits
and hospitalizations [@fried01a], and long-term care admissions
[@pajewski19; @rockwood06; @rockwood05]. The Fried phenotype
[@fried01a] and Rockwood deficit accumulation index
[@rockwood07; @mitnitski01a], are the most commonly used methods for
operationalizing frailty. There is reasonable convergence between these
two approaches [@li15; @malmstrom14] but the deficit accumulation
approach does not require individual questionnaires nor physical
assessments yet offers similar [@malmstrom14] or possibly better
[@kulminski08] predictive accuracy.

Possible points still left to cover:

-   outcomes important to clinical care of older patients

-   how care could be improved with better modeling of these outcomes

-   information about our site


# Methods

## Population

A random 1% sample (N=14,844) was drawn from the deidentified patient
records of a large academic health center and its teaching hospital
partner. Visits during which patient age was less than 18 years old were
excluded and then patients who had fewer than three visits in the
remaining data were excluded. To avoid bias, for each patient an index
visit date was chosen and only data recorded on or after that date was
used in analysis. To avoid distorted results in patients with sparse
visit histories, those who had fewer than two visit-dates after index
visit assignment were removed from the sample as were patients whose EFI
was never higher than 0. Finally, the patients were randomly assigned to
a development cohort (N=2,497 patients, 52,372 visit-days) or a testing
cohort (N=3,220 patients, 56,320 visit-days). Sensitivity analysis was
done to see the effect of leaving in data from all visits by adult
patients and the overall direction of EFI's effect was the same but
[the performance improvement relative to patient age was inflated].
All decisions about data processing and statistical analysis were made
using only the development cohort and blinded to the testing cohort. For
publication, the same analysis scripts were run on the testing cohort
and used to create all results reported here (the development version of
each table and figure is available in the supplemental materials). The
baseline characteristics of the testing cohort are shown in Table
\@ref(tab:table1).

***

```{r table1header}
    pander(rbind(c('.','','','','','','',''))
           ,caption='(\\#tab:table1) Cohort demographics\n\n');
```


```{r table1,results='asis'}
if(.outfmt == 'html'){
  .tb1formula <- setdiff(names(dat04),c('language_cd','Frailty Stage'
                                        ,'Median Frailty','patient_num'
                                        ,'age_at_death_days'
                                        ,'age_at_visit_days'
                                        ,setdiff(v(c_response)
                                                 ,.outcomenames))) %>%
    paste0('`',.,'`',collapse='+') %>% paste('~',.,'|`Frailty Stage`') %>%
    formula;
  
  tb1 <- table1(.tb1formula,data=dat04) %>%
    submulti(dct0[,c('colname','dispname')],'partial');
  
  print(tb1);
};
```


## Electronic Frailty Index

@clegg16a developed an electronic frailty index (EFI) for UK health systems
following the methodology of Mitnitsky, Rockwood, et al.
[@mitnitski01a; @song10; @searle08a]. Recently, @pajewski19 adapted the EFI
to use ICD10 and ICD9. We further built on this work by also mapping
laboratory tests to specific LOINC codes. For each patient visit, all
distinct diagnoses and abnormal lab results over the preceding two-year
window for that patient were aggregated into a single EFI, a numeric value
that can range from 0 to 1 (but in practice seldom exceeded 0.6). As
mentioned above, we omitted all visits prior to a randomly select index
visit for each patient. This did not interfere with EFI calculation because
those EFI values were calculated separately for every distinct patient-date
in our health system, and then joined to the EHR data.

## Outcomes

The primary outcomes we predicted with EFI were falls (any ICD9 codes in
E880-E888.9, E987.x, V00.x, or ICD10 codes in W00-W19), hospital
admissions (EHR encounter type = 'inpatient'), hospital-acquired
infections (having any ICD10 code associated with hospital-acquired
infections per Southern et al. @southern17 ), discharge to
intermediate care facilities (ICF) or skilled nursing facilities (SNF)
(using discharge disposition codes, for patients whose admit-source for
that encounter was 'Home'), and all-cause mortality insofar as it can be
ascertained from the EHR discharge disposition or vital status (which,
in the absence of linkage to external death indexes are likely to
under-report the actual mortality and so should be interpreted with
caution).


## Statistical Analysis

For each outcome of interest, we used a Cox proportional hazard model to
estimate the risk of the first occurrences of the outcome after the
patients' respective index visits using EFI as the predictor. Unlike
earlier studies, we treated EFI as a time-varying numeric predictor with
multiple followups per patient. Since only the first occurrence was
being predicted, a recurring event model was not necessary. As a
comparison, for each outcome a second analysis was done that was
identical to the first except that patient age at visit (in days) was
used as the predictor.

# Results

Table \@ref(tab:tb2) shows the results of Cox proportional hazard models for
each of the responses, with EFI as the predictor. For each 0.1 increase in
EFI, we found at least a doubling of risk: `r pander(snip_resultsfold)`.
Furthermore, the probability of delayed discharge increased and thus longer
stays were observed in patients with an EFI > 0.19.
The p-values shown have been adjusted for multiple comparisons
(`r as.english(nrow(tb2)) %>% as.character` outcomes reported in
one study) using the @holm79 method and in all cases are highly
significant. It is well established in gerontology that an enormous and
diverse range of poor health outcomes are correlated with an individual's
age, but chronological age provides no information about the progression of
biological age in an individual. It has long been a priority in aging
research to find a metric for aging that is subject to individual variation
and intervention. Rockford et al. have pointed out {ref} that deficit
accumulation has the hallmarks of such a measure.


```{r tb2}

tb2[,c('predictor','Outcome','β^ (95% CI)','fold-change (95% CI)','SE','Z','P, adjusted')] %>%
  pander(digits=3, row.names=FALSE,caption='(\\#tab:tb2) Cox-proportional hazards with EFI as a predictor');
```


Therefore, we compared the predictive accuracy of the EFI to that of
patient age. For each outcome, we fit an additional Cox proportional
hazard model, this time using age at visit as the predictor variable. It
appears, at least in this population, that not only can EFI accurately
predict mortality and poor patient outcomes without any knowledge of
patient age but it does so better than patient age.


```{r tb3}
pander(tb3,caption='(\\#tab:tb3) Performance of EFI models');
```


In Table \@ref(tab:tb3) we compare the performance of EFI and age as
predictors for each of the `r as.english(nrow(tb2)) %>% as.character`
outcomes. In all cases, the EFI models has a robustly lower AIC and
log-likelihood than the respective patient-age models, as well as a higher
concordance (in all cases greater than 0.7). Cox models incorporating EFI
together with age, with or without an interaction term, did not have a
significantly better fit than the univariate EFI model (not shown, available
in supplement).

In figures \@ref(fig:kmplots) a-`r letters[length(fits)]` we show Kaplan-Meier 
plots for each of the outcomes stratified by whether EFI is greater than 0.19 
(Frail=TRUE) or less (Frail=FALSE) [@stow18].


```{r kmplots, fig.show="hold", fig.height=4, fig.width=12, fig.cap=" - "}
# . kmplots ----
.subfig <- 0;
# for(jj in fits) {
#   print(jj$plot + labs(caption=paste0('Figure 1 ',letters[.subfig],'. \n')) +
#           theme(plot.caption=element_text(size=15)) );
# }
#panderOptions('knitr.auto.asis', FALSE);
for(jj in fits) {.subfig <- .subfig + 1;
with(jj,{
  message(dispname);
  print(ggmatrix(multiplot,nrow=1,ncol=3,legend=grab_legend(plot),title=dispname
           ,xAxisLabels = names(multidata)
           ,xlab=plot$label$x
           ,ylab=plot$label$y) + 
          labs(caption=paste0('Figure 1 ',letters[.subfig],'. \n')));
  cat('\n\n\n');
  # pander(modelsummary);
  # cat("\n******\n");
  })};

```



# Discussion

Our data shares the fundamental limitation of the EHR system from which
it was obtained: like all EHR systems, it only has information that
providers and coders put into it. Events taking place outside the health
system or at un-connected health systems are not visible to our analysis.
On the other hand, providers who rely on EHR systems at point of care
are also working under these limitations on. The data we used is
representative of this scenario, and despite the limitations EFI
provides accurate predictions of poor patient outcomes. Because our
implementation of the Rockwood index real EHR data, it is more directly
transferable to clinical use than implementations based on curated
registries. This suggests that EFI is most accurate for patients who
have accumulated a reasonable in-system visit-history. Further work is
needed to find a more precise relationship between the length of a
patient's visit-history and the accuracy of EFI and better distinguish
genuinely non-frail patients from those who get most of their care
outside the researchers' health system.

Vital status is often out of date in EHR systems if the patient did not
die in the hospital or shortly after their visit, so our data may
under-represent the true mortality rates. Nevertheless, the relationship
between EFI and mortality risk observed here agrees with previous EFI
studies. Unlike previous studies we sampled all adults rather than just
older adults so we could assess the performance of EFI across the
lifespan and include individuals who might have early-onset frailty.

There is no gold standard method to assess Frailty in clinical practice.
Currently available frailty assessment tools used in geriatric practice have
good validity (for example @fried01a) but these are time
intensive and often difficult to implement in a busy general practice.
Assessing frailty helps clinicians identify high risk patients and tailor
interventions to prevent health decline and poor outcomes. Because of the
simplicity of the Rockwood Frailty Index, it is more likely to be adopted by
clinicians in a busy practice.

Other possible points to cover:

-   What we learned by not restricting the sample to older adults (or
    move to methods/theory paper?)

-   What agrees and what disagrees with previous work?
-   What is new?
    -   Per-visit EFI allows plotting an EFI trajectory over time
    -   Strictly limiting data elements to those we can count on being
        available in any EHR system, at any site
-   future work
-   theoretical and practical implications



# Conclusions

TBD

# Acknowledgments

This work was supported by NIH/NCATS UL1TR001120 (AFB, KRS), the Long School
of Medicine KL2 Award (AFB), ...



# References



---
date: '2020-09-22'

---
