---
title: "Development and validation of an EHR-based frailty index for prediction patient safety events in a health system."
#' author: 'Alex Bokov ^1,✉^, Sara Espinoza ^1^, Chandana Tripathy ^1^, and Kathleen R. Stevens ^1^'
css: "production.css"
abstract: | 
  We conducted a retrospective cohort study on structured data of XXXX patients
  randomly sampled from all adult patients in the Epic EHR system of a large 
  South Texas academic health center and its teaching hospital partner to 
  determine whether our implementation of the Rockwood frailty index (eFI) can 
  predict iatrogenic patient safety events using only codes and numeric values 
  routinely collected in EHR systems. According to both the c-statistic and the 
  AIC of the Cox proportional hazard models we used to analyze the data, eFI 
  performs well as a predictor of infections, non-operative trauma, cardiac 
  complications, and of coded events categorized as patient safety indicators 
  (PSI).

documentclass: article
description: 'Manuscript'
clean: false
bibliography: efi_paper.bib
csl: harvard-cite-them-right.csl
output:
  bookdown::html_document2:
    fig_caption: true
    self_contained: true
    number_sections: false
    keep_md: true
    clean: false
  bookdown::word_document2:
    reference_docx: efi_template.docx
    fig_caption: true
    self_contained: true
    keep_md: true
    number_sections: false
---

```{r load_deps, echo=FALSE, message=FALSE, warning=FALSE,results='hide'}
.projpackages <- c('GGally','pander','dplyr','ggplot2','data.table'
                   ,'survival','broom','forcats','table1','english','bookdown'
                   ,'DiagrammeR','DiagrammeRsvg','rsvg');
.deps <- c( 'analysis.R' );
.debug <- 0;
message(paste0('\n\n working dir: ',getwd(),'\n\n'));
cat('\n\nAvailable Files:',list.files(),sep='\n','\n');
.junk<-capture.output(source(file.path('./scripts/global.R'),chdir=TRUE,echo=!1
                             ,local=TRUE));
# Set some formatting options for this document
panderOptions('table.alignment.default','right');
panderOptions('table.alignment.rownames','right');
panderOptions('table.split.table',Inf);
panderOptions('p.wrap','');
panderOptions('p.copula',', and ');
panderOptions('graph.fontfamily','serif');
# theme_get() gives default theme
theme_set(theme_bw(base_family = 'serif',base_size=14) +
            theme(strip.background = element_rect(fill=NA,color=NA)
                  ,strip.text = element_text(size=15)));
knitr::opts_chunk$set(echo=.debug>0, warning=.debug>0, message=.debug>0);
knitr::knit_hooks$set(inline = function(xx) {prettyNum(xx, big.mark=",")});


# detect current output format
.outfmt <- knitr::opts_knit$get('rmarkdown.pandoc.to');
if(is.null(.outfmt)){
  .outfmt <- if(knitr::is_html_output()) 'html' else {
    if(knitr::is_latex_output()) 'latex' else {
      if(interactive()) 'html' else {
        'unknown';
    }}}};
message('.outfmt = ',.outfmt);

.currentscript <- current_scriptname('efi_patientsafety.Rmd');
.outcomenames <- c('vi_c_psi','vi_c_hospinfect','vi_c_hosptrauma'
                   ,'vi_c_cardiac');
snip_outcomes <- rname(.outcomenames);
snip_todates <- dat03[,.(xx=max(start_date)),by=patient_num] %>%
  with(range(xx)) %>% paste0(collapse=' and ');
snip_fromdates <- dat03[,.(xx=min(start_date)),by=patient_num] %>%
  with(range(xx)) %>% paste0(collapse=' and ');

# limit the data to items of interest


# snippets ----

```
# Introduction/Background

*[TRIPOD 3.a] Explain the medical context (including whether diagnostic or prognostic) and rationale for developing or validating the multivariable prediction model, including references to existing models.*
*[TRIPOD 3.b] Specify the objectives, including whether the study describes the development or validation of the model or both.*

Frailty is the lifelong erosion of stress resistance and accumulation of
impairments across multiple physiological systems. Among older community
dwelling adults 32% have been classified as pre-frail and 24% have been
classified as frail [@hoover13]. Frailty predicts disability, injurious falls,
and mortality [@pajewski19], emergency room visits and hospitalizations
[@fried01a], and long-term care admissions [@pajewski19; @rockwood06;
@rockwood05]. The Fried phenotype [@fried01a] and Rockwood deficit accumulation
index [@rockwood07; @mitnitski01a], are two commonly used measures of frailty. 
There is reasonable convergence between these two approaches 
[@li15; @malmstrom14] but the former is better suited to large-scale deployment
in EHR systems because it does not require patient questionnaires nor 
dedicated physical assessments. Furthermore, Kulminsky @kulminski08 found 
evidence for the index outperforming the phenotype. 

Our version of the electronic frailty index (eFI) improves on previous EHR 
adaptations of the Rockwood index [@clegg16a; @pajewski19] by eliminating the 
need for manually-curated lab reference ranges and by calculating the eFI over
a rolling two-year window for every available patient encounter rather than at 
one fixed time point. The scripts and usage notes are freely available online.

Improving and maintaining patient safety is a key goal for every health system. 
A simple and rapid frailty assessment would help clinicians can tailor 
interventions to avoid poor outcomes in their patients. Here our objective was 
to develop and validate (against hold-out data) eFI as a predictor of 
preventable infections, trauma, cardiac complications, and patient safety events
in general.


# Methods

## Source of data

*[TRIPOD 4.a] Describe the study design or source of data (e.g., randomized trial, cohort, or registry data), separately for the development and validation data sets, if applicable.*

*[TRIPOD 4.b] Specify the key study dates, including start of accrual; end of accrual; and, if applicable, end of follow-up. *

A random 1% sample (`r subset(consort,node=='start')$patients` patients, 
`r subset(consort,node=='start')$patdays` visit-days) was drawn from 
deidentified i2b2 data warehouse records representing the entire patient 
population in the EHR system of a large academic health center and its teaching 
hospital partner. 

The nominal date of index encounters ranged between `r snip_fromdates` while 
that of final encounters ranged between `r snip_todates`. The actual
dates were within one year of these dates due to deidentification.

## Participants

*[TRIPOD 5.a] Specify key elements of the study setting (e.g., primary care, secondary care, general population) including number and location of centres.*
*[TRIPOD 5.b] Describe eligibility criteria for participants. *
*[TRIPOD 5.c] Give details of treatments received, if relevant. *

The patient-visits in the data are a representative sample of the mix of all 
primary care, specialist, and inpatient encounters of UT Health San Antonio and 
the University Health System. As summarized in Figure \@ref(fig:consort), 
encounters during which the patient was below the age of 18 were excluded and
patients who had fewer than three such encounters were excluded. To avoid the
bias that would accompany always using the first valid visit as a starting 
point, an index date was randomly selected for each patient from among all 
their visit dates and only data recorded on or after that visit was used in 
analysis (not counting the eFI values which were pre-calculated over for every 
patient-visit in the EHR system, obtained as a separate file, and linked to 
the main dataset). After this step, patients who had fewer than two visits 
remaining were also excluded.

Some patients whose eFI never goes above 0 are non-frail, but there is no 
reliable way to distinguish them from those who have a 0 eFI for artifactual 
causes-- e.g. incomplete records, sparse encounter histories. Furthermore, at 
the tail ends of their histories most patients had date-only records during 
which eFI was 0 and no diagnoses, procedures, nor lab results were recorded. 
These may have been missed appointments or accesses to patient files without 
actual encounters. We therefore excluded patients who never had 
an eFI greater than 0 and for those that remained we excluded all visits 
after the last procedure, diagnosis, lab, or eFI (whichever happened last) was recorded.

Finally, the patients were split into development (
`r subset(consort,node=='dev')$patients` patients, 
`r subset(consort,node=='dev')$patdays` visit-days) and validation (
`r subset(consort,node=='test')$patients` patients, 
`r subset(consort,node=='test')$patdays` visit-days) subsets.

```{r consort,fig.cap='CONSORT Diagram'}
grViz(consortgv);
```
This study was performed using only retrospective EHR data with no 
treatment, intervention or any other interaction with patients. The exposure 
groups were determined by patient eFI as will be discussed below.

## Sample size

*[TRIPOD 8] Explain how the study size was arrived at.*

We limited ourselves to a 1% initial sample for two reasons. First, given 
anticipated utility of the eFI for predicting other clinically important
outcomes in future studies, and given that our approach uses information from a 
large time-span for each sampled patient, we needed to avoid depleting the un-analyzed patient population in these early stages. Secondly, we needed to 
stay within processing-time and file-size constraints of our research
data warehouse team. Extrapolating from the effect sizes in @pajewski19, events 
with an overall prevalence of 5% and a hazard ratio of 1.33 between frail and 
non-frail groups should be detectable with a power well over 0.80. 

## Outcomes and Predictors
*[TRIPOD 6.a] Clearly define the outcome that is predicted by the prediction model, including how and when assessed. *

*[TRIPOD 6.b] Report any actions to blind assessment of the outcome to be predicted. *
*[TRIPOD 7.b] Report any actions to blind assessment of predictors for the outcome and other predictors. *

We drew on the expertise of the clinicians on our team (CT and SE) to select 
outcomes that are meaningful, relevant, and useful in guiding clinical care 
planning. These outcomes were: `r pander(wrap(snip_outcomes,'*'))`. To assess 
these outcomes in EHR data we used ICD-10 code mappings developed by 
@southern17. 

Insofar that any other decisions about processing and statistical analysis 
required data, we used only the development data and were blinded to the 
validation data. After the analysis scripts were finalized for publication, 
they were re-run on the validation data and used to generate all the results 
reported here. The development version of each table and figure is available in the 
supplemental materials. 

*[TRIPOD 7.a] Clearly define all predictors used in developing or validating the multivariable prediction model, including how and when they were measured.*

Predictors (i.e. deficits on which eFI is calculated) were chosen prior to 
extraction of the research data, based on the frailty index implemented by 
@clegg16a for UK health systems and then adapted to ICD10 by @pajewski19. The
underlying methodology was developed by Mitnitsky, Rockwood, et al. [@mitnitski01a;@song10; @searle08a]. We built upon their work by mapping 
laboratory tests to specific LOINC codes and omitting data elements which are 
not universally available in EHR systems. We made the mapping tables and sample 
SQL code available for public use and discussion [@bokov20]. For each patient 
visit, all distinct diagnoses and abnormal lab results over the preceding 
two-year window were aggregated into a single eFI-- a numeric value that
can range from 0 to 1 (but in practice seldom exceeds 0.6). Though we omitted 
all visits prior to a randomly select index visit for each patient (see 
[Participants] section, above) this did not interfere with EFI calculation 
because eFI values were calculated en-masse for every distinct patient-date in 
our health system and then matching records were joined to the study sample.


## Missing data

*[TRIPOD 9] Describe how missing data were handled (e.g., complete-case analysis, single imputation, multiple imputation) with details of any imputation method. *

Our missing data strategy approximates a complete-case analysis in 
that we excluded patients with too few visits or missing eFI. No imputation was
done on the remaining records; absence of ICD10 codes for the respective 
outcomes was interpreted as non-occurrence. This is in line with how discrete
codes are used in the EHR.

## Statistical analysis methods

*[TRIPOD 10.a, development only] Describe how predictors were handled in the analyses. *

*[TRIPOD 10.b, development only] Specify type of model, all model-building procedures (including any predictor selection), and method for internal validation.*

*[TRIPOD 10.d] Specify all measures used to assess model performance and, if relevant, to compare multiple models. *

For each outcome of interest, we used Cox proportional hazard models [@cox72]
with using eFI as the predictor to estimate the risk of the first occurrences of
the respective outcomes after the patients' index visits. Unlike earlier 
studies, we treated eFI as a time-varying numeric predictor with multiple 
follow-ups per patient. Since patients were followed up to the first occurrences 
the standard form of the Cox model was sufficient. Harrell's c-statistic 
[@harrell84; @harrell96] was used to assess model performance 
(Table \@ref(tab:tb3)).

*[TRIPOD 10.c, validation only] For validation, describe how the predictions were calculated. *


*[TRIPOD 10.e, validation only] Describe any model updating (e.g., recalibration) arising from the validation, if done.*

No updates or model recalibration were done after running the analysis on the 
validation data.

## Risk groups

*[TRIPOD 11] Provide details on how risk groups were created, if done. *


## Development vs. validation

*[TRIPOD 12, validation only] For validation, identify any differences from the development data in setting, eligibility criteria, outcome, and predictors. *

There were no differences in the choice of predictors, outcome, inclusion 
criteria, or clinical setting between the development and validation data 
because both were mutually exclusive, randomly-assigned subgroups of patients 
from the same sample.

# Results

## Participants

*[TRIPOD 13.a] Describe the flow of participants through the study, including the number of participants with and without the outcome and, if applicable, a summary of the follow-up time. A diagram may be helpful. *

*[TRIPOD 13.b] Describe the characteristics of the participants (basic demographics, clinical features, available predictors), including the number of participants with missing data for predictors and outcome. *

*[TRIPOD 13.c, validation only] For validation, show a comparison with the development data of the distribution of important variables (demographics, predictors and outcome). *

## Model development 

*[TRIPOD 14.a, development only] Specify the number of participants and outcome events in each analysis. *

*[TRIPOD 14.b, development only] If done, report the unadjusted association between each candidate predictor and outcome.*

## Model specification

*[TRIPOD 15.a, development only] Present the full prediction model to allow predictions for individuals (i.e., all regression coefficients, and model intercept or baseline survival at a given time point).*

*[TRIPOD 15.b, development only] Explain how to the use the prediction model.*

## Model performance

*[TRIPOD 16] Report performance measures (with CIs) for the prediction model.*
```{r tb3}
subset(tb3,Outcome %in% .outcomenames & Predictor == 'Frailty') %>% 
  mutate(Outcome = rnameshort(Outcome)
         ,`Concordance (CI)`=paste0(round(Concordance,3)
                                    ,' ('
                                    ,round(Concordance-1.96*`SE Concordance`,3)
                                    ,', '
                                    ,round(Concordance+1.96*`SE Concordance`,3)
                                    ,')')) %>% 
  select('Outcome','Concordance (CI)','Log Likelihood') %>% 
  pander(caption='(\\#tab:tb3) Performance of EFI models',row.names=F);

```


## Model-updating

*[TRIPOD 17, validation only] If done, report the results from any model updating (i.e., model specification, model performance).*

# Discussion

## Limitations

*[TRIPOD 18] Discuss any limitations of the study (such as nonrepresentative sample, few events per predictor, missing data). *

## Interpretation

*[TRIPOD 19.a, validation only] For validation, discuss the results with reference to performance in the development data, and any other validation data. *

*[TRIPOD 19.b] Give an overall interpretation of the results, considering objectives, limitations, results from similar studies, and other relevant evidence. *

## Implications

*[TRIPOD 20] Discuss the potential clinical use of the model and implications for future research. *

# Conclusions

## Supplementary information

*[TRIPOD 21] Provide information about the availability of supplementary resources, such as study protocol, Web calculator, and data sets. *

# Acknowledgments

*[TRIPOD 22] Give the source of funding and the role of the funders for the present study. *


# References



---
date: '2021-08-08'

---
